On selection of a coupling method to cope with change of priors in one-vs-one classification

### 2025-06-12 experiment.tex
Revision

\subsection{Experiment 1: Accuracy of parameter-less coupling methods on benchmark datasets} \label{sec:exp1}

@@

From Figure \ref{fig:exp1-plot1} we can clearly see that Wu-Lin-Weng's method has the best performance. On some datasets it trounces normal coupling method by a wide margin. The edge over radial method is smaller, with the exception of usps dataset where its accuracy is higher by remarkable 30\%.

\begin{figure}
	\includegraphics{graphs/exp1-plot1.pdf}
	\caption{0-1 score plotted against the logarithm of $tol$ hyper-parameter, evaluated for different datasets.}
	\label{fig:exp1-plot1}
\end{figure}

\subsection{Experiment 2: Accuracy of Bayes covariant methods with parameters}
 \label{sec:exp2}

The method of Wu-Lin-Weng provides leading performance among the \emph{non-parametric} methods examined in the previous experiment. However in view of Theorem \ref{thm:K3} the Wu-Lin-Weng's method is not Bayes covariant.

In this experiment we investigated whether we can (always) find  a (possibly \emph{parametric}) Bayes covariant coupling method that could match the performance of Wu-Lin-Weng's method. We investigate  the case of three class datasets. This is a natural starting point, since it is the smallest number of classes where the question is meaningful to asks.  Our experiment has three parts.

In the first part, we consider whether four parameterless Bayes covariant method could match the performance of Wu-Lin-Weng method. In our comparison we include the normal method, as well as the three arboreal methods $ \boldsymbol{s}_1,~ \boldsymbol{s}_2,~ \boldsymbol{s}_3$. We consider all possible triples of classes in all datasets. The resulting statistics are shown in Table \ref{tab:step2}.

@@

We can see that in slightly less than half of the cases none of the four parameterless Bayes covariant methods was able to match the performance of Wu-Lin-Weng's method. Therefore we will extend search to Bayes covariant methods with trainable parameters. Let us define two parameter family of Bayes covariant coupling methods
\begin{align}
a_1 \boldsymbol{s}_1 \oplus a_2 \boldsymbol{s}_2 \oplus a_3 \boldsymbol{s}_3,\quad\textrm{where } a_1 + a_2 +a_3 = 1 \label{eq:family}
\end{align}
We note that the parameter space of this family (the hyperplane $a_1 + a_2 + a_3 =1$) is not compact.

In the second part of our experiment we  conduct a grid search for methods matching performance of Wu-Lin-Weng method. We examine methods for which the coefficients $a_i$ are nonnegative $a_i \geq 0$.  To reduce computational burden we selected 1000 random chosen three-class subproblems from those where Wu-Lin-Weng's method outperformed normal coupling. The summary statistics are shown in Figure \ref{fig:par-bc}.

\begin{figure}[!ht]
\includegraphics{graphs/exp2-summary.pdf}
\caption{Results of grid search for parametric Bayes covariant methods matching performance of Wu-Lin-Weng's method}
\label{fig:par-bc}
\end{figure}

We can see that only in minority of cases we found a parametric method matching the performance of Wu-Lin-Weng's. This finding comes with caveat that we searched only in the simplex $a_i \geq 0$.

Therefore in the third part of our experiment  we looked whether a parametric Bayes covariant method in family \eqref{eq:family} for parameters \emph{outside} of this simplex could perform better. We conducted another search but this time
\begin{itemize}
	\item We searched within a much larger rectangle $D$ containing the simplex $a_i\geq 0$, which in this case is a triangle $T$.
	\item We used Brier score as a smooth proxy for measuring the quality of multi-class classifier. This score is much smoother than 0-1 score which measures accuracy.
\end{itemize}

>>>>

\subsection{Experiment: Accuracy of canonical coupling methods on benchmark datasets} \label{sec:exp1}

@@

From Figure \ref{fig:exp1-plot1} we can clearly see that Wu-Lin-Weng's method has the best performance. On some datasets it trounces normal coupling method by a wide margin. The edge over the radial method is smaller with only mnist and usps datasets showing a significant difference for smaller values of $tol$ hyperparameter.

\begin{figure}
	\includegraphics{graphs/exp1-plot1.pdf}
	\caption{Success rate plotted against the logarithm of $tol$ hyper-parameter, evaluated for different datasets.}
	\label{fig:exp1-plot1}
\end{figure}

\subsection{Experiment: Accuracy of Bayes covariant methods without parameters}
 \label{sec:exp2}

The method of Wu-Lin-Weng provides leading performance among the canonical {nonparametric} methods examined in the previous experiment. However in view of Theorem \ref{thm:K3} the Wu-Lin-Weng's method is not Bayes covariant.

In this experiment we investigated whether we can (always) find  a (possibly \emph{parametric}) Bayes covariant coupling method that could match the performance of Wu-Lin-Weng's method. We investigate  the case of three class datasets. This is a natural starting point, since it is the smallest number of classes where the question is meaningful to asks. All experiments use value of 0.05 for $tol$ hyperparameter based on results of Section \ref{sec:exp1}. Our experiment has three parts.

In the first part, we consider whether four parameterless Bayes covariant method could match the performance of Wu-Lin-Weng method. In our comparison we include the normal method, as well as the three arboreal methods $ \boldsymbol{s}_1,~ \boldsymbol{s}_2,~ \boldsymbol{s}_3$ that are not canonical. We consider all possible triples of classes in all datasets. The resulting statistics are shown in Table \ref{tab:step2}.

@@

We can see that in slightly less than half of the cases none of the four parameterless Bayes covariant methods was able to match the performance of Wu-Lin-Weng's method. Therefore we will extend search to Bayes covariant methods with trainable parameters.

\subsection{Experiment: A parametric family}

Let us define two parameter family of Bayes covariant coupling methods
\begin{align}
a_1 \boldsymbol{s}_1 \oplus a_2 \boldsymbol{s}_2 \oplus a_3 \boldsymbol{s}_3,\quad\textrm{where } a_1 + a_2 +a_3 = 1. \label{eq:family}
\end{align}
Note that this family includes the normal coupling method as well as the three arboreal methods $\boldsymbol{s}_1, \boldsymbol{s}_2, \boldsymbol{s}_3$.

In this experiment we  conduct a grid search for methods matching performance of Wu-Lin-Weng method. Grid search works best in low dimension, when restricted to a compact subset. To that end we restrict our search to the simplex formed with nonnegative coefficients $a_i \geq 0$, which form a compact subset.  To reduce computational burden we selected 1000 random chosen three-class subproblems from those where Wu-Lin-Weng's method outperformed normal coupling. The summary statistics are shown in Figure \ref{fig:par-bc}.

\begin{figure}[!ht]
\includegraphics{graphs/exp2-summary.pdf}
\caption{Comparison of success rate for parametric Bayes covariant methods \eqref{eq:family} vs. Wu-Lin-Weng's method}
\label{fig:par-bc}
\end{figure}

We can see that only in minority of cases we found a parametric Bayes covariant coupling method matching the performance of Wu-Lin-Weng's method. This finding comes with caveat that we searched only in the simplex $a_i \geq 0$.

\subsection{Example: }

We determined that among the 1000 three-class problems we have examined, the largest gap between Bayes-covariant parametric methods and Wu-Lin-Weng method occurred for mnist dataset and classes 5,6,9. Figure

\begin{itemize}
	\item We searched within a much larger rectangle $D$ containing the simplex $a_i\geq 0$, which in this case is a triangle $T$.
	\item We used Brier score as a smooth proxy for measuring the quality of multi-class classifier. This score is much smoother than 0-1 score which measures accuracy.
\end{itemize}
